Graded Call-By-Push-Value

### 2020-07-10 popl21/popl21.tex
[ popl21 ] submission

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Call-by-push-value (CBPV) is a simply typed lambda calculus that polarizes
types into value and computation types and can thus express both
call-by-name and call-by-value evaluation in the presence of effects.
Semantically, effects are modeled by a monad, and computation types as
algebras over this monad.
%
Effect type systems usually express more information than the presence
of an effect; often effects are categorized by preordered monoid
where the monoid operation represents accumulation of effects and the
order expresses effect subsumption, in analogy to subtyping.
In this work, a \emph{graded} version of CBPV is presented where the typing
of computations likens effect typing.  Semantically, computation types
are then represented as graded monad algebras.
%
Observing that the value types of CBPV can be interpreted as comonad coalgebras, we further present a version of CBPV that has coeffects graded by a preordered semiring.  Value types and contexts are interpreted as graded comonad coalgebras, allowing resource-aware interpretations of CBPV.
%
Finally, we combine the two systems into a fully graded version of CBPV where both effects and coeffects are graded.  Thus turns out to be possible without specifying an interaction between effects and coeffects.
\end{abstract}

@@

CBPV polarizes types into value types $P$ and computation types $N$
where the former embed into the latter as $\diamond P$ by the formal
``monad''~$\diamond$ and the latter embed into the former as $\Box N$
by ``thunking''~$\Box$.  In Levy's \emph{behavioral semantics},
functor $\diamond$ is a left adjoint to functor $\Box$ and need not be
a monad.  Quite the opposite, when modelling store
$\diamond P = P \times S$ is a comonad and $\Box N = S \to N$ a monad.
Contrastingly, in the above explained \emph{algebra semantics},
$\diamond$ is indeed modeled by a monad $\T$, yet $\Box$ just by the
identity.  It is however possible to model $\Box$ by a monoidal
comonad $\D$, and positive types by comonad coalgebras with an action
$\texpose : P \to \D\,P$ that makes the services of the comonad
available at all value types, not just at $\Box N$.

Building on these observations, we investigate in this article how
CBPV %would look like if it would
can
keep track of not just the presence
of an effect or coeffect but also the kind of effect and coeffect.
Semantically, precise information about effects can be obtained by
using a family of monads $\T[e]$ graded over effect classifiers $e$
drawn from a preordered monoid \cite{katsumata:popl14}.  For CBPV, we
generalize the concept of monad algebras $N$ to \emph{graded} such
ones, $N_e$.  Syntactically, computation typing $\CompTy \Gamma t e N$
is extended by an effect classifier $e$
(as usual in type and effect systems \citep{nielsen:effect99})
such that a computation can be
interpreted by a morphism $\Gamma \to N_e$.

Coeffect type systems
\citep{brunel:esop14,orchard:icfp14,ghicaSmith:esop14} have been
developed to track resource consumption by attaching usage information
to each variable in the typing context.  Coeffect typing generalizes
linear typing \citep{girard:linear} to quantitative typing
\citep{sergeyVytiniotisPeytonJones:popl14,mcBride:wadler60,atkey:lics18}
and subsumes sensitivity analysis \citep{reedPierce:icfp10} and static
information control flow \citep{volpano:jcs96} aka security typing
\citep{abadiBanerjeeHeintzeRiecke:popl99}.
Coeffects can be modeled by a comonad $\D[r]$ graded over resource
qualifiers $r$ drawn from a preordered semiring.  For the sake of
coeffect-graded CBPV, we generalize graded comonads to graded comonad
coalgebras.  Value types $P$ and contexts $\Gamma$ are then
interpreted as such coalgebras.  Value typing
$\qValTy \gamma \Gamma v P$ and computation typing
$\qCompTy \gamma \Gamma t N$ are based on linear typing an come
equipped with a resource context~$\gamma$.  Semantically, values are
interpreted as morphisms $\Gamma_{r\gamma} \to P_r$ allowing the
``multiplication'' or \emph{scaling} of a value by $r$ % natural in $r$
and computations as morphisms $\Gamma_\gamma \to N$.

Finally, effect and coeffect graded CBPV can be combined into a fully
graded CBPV calculus, surprisingly without sorting out any interaction
between effects and coeffects, such as the distributive laws of
\citet{orchard:icfp16}.  We credit the smoothness of the integration
to the careful placement of monad and comonad in CBPV's type system,
and the lack of need for scaling % of not just values but also
of computations.

@@

@@

\citet{mcDermottMycroft:esop19} observe that CBPV can represent
call-by-name and call-by-value evaluation strategies but not
call-by-need.  As a remedy, they extend CBPV to ECBPV by variables
$\underline x : \diamond P$ that hold monadic values, plus a
corresponding let-binding construct.  They further present an
effect-graded version of ECBPV, drawing effect qualifiers from a
preordered monoid.  In contrast to our version, effects annotations
$f$ are stored at monadic types $\compty f P$.  Instead of our
judgement $\CompTy \Gamma t e N$ that accumulates effect traces $e$
and a semantic interpretation of computation types $N$ as graded monad
algebras, they define a syntactic operation $\compty f N$ that pushes
down effect qualifier $f$ to the monadic types $\compty f' P$ in $N$
where the effects are combined to $\compty{f \bu f'} P$.  This way of
handling effect grading seems equivalent to our approach, albeit it is
syntactic rather than semantic.

In previous work, \citet{mcDermottMycroft:ocs18} use coeffect typing
to track variable usage in call-by-need simply-typed lambda calculus
with conditionals.  Compared to the usual resource semiring $\R$
tailored to call-by-name usage analysis, their coeffect algebra is
more expressive: it can produce traces of variable uses, including
bound variables.  They show how graded effect tracking can be
simulated by the coeffect typing, \eg, for the effect of
non-determinism.  Naturally, they focus on operational semantics.

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Call-by-push-value (CBPV) is a simply typed lambda calculus that
polarizes types into value and computation types and can thus
express both call-by-name and call-by-value evaluation in the
presence of effects.  Semantically, effects are modeled by a monad,
and computation types as algebras over this monad.

Effect type systems usually express more information than the
presence of an effect; often effects are categorized by preordered
monoid where the monoid operation represents accumulation of effects
and the order expresses effect subsumption, in analogy to subtyping.
In this work, a \emph{graded} version of CBPV is presented where the
typing of computations likens effect typing.  Semantically,
computation types are then represented as graded monad algebras.

Observing that the value types of CBPV can be interpreted as comonad
coalgebras, we further present a version of CBPV that has coeffects
graded by a preordered semiring.  Value types and contexts are
interpreted as graded comonad coalgebras, allowing resource-aware
interpretations of CBPV.

Finally, we combine the two systems into a fully graded version of
CBPV where both effects and coeffects are graded.  This turns out to
be possible without specifying any interaction between effects and
coeffects.
\end{abstract}

@@

CBPV polarizes types into value types $P$ and computation types $N$
where the former embed into the latter as $\diamond P$ by the formal
``monad''~$\diamond$ and the latter embed into the former as $\Box N$
by ``thunking''~$\Box$.  In Levy's \emph{behavioral semantics},
functor $\diamond$ is a left adjoint to functor $\Box$ and need not be
a monad.  Quite the opposite: when modelling store,
$\diamond P = P \times S$ is a comonad and $\Box N = S \to N$ a monad.
Contrastingly, in the above explained \emph{algebra semantics},
$\diamond$ is indeed modeled by a monad $\T$, yet $\Box$ just by the
identity.  It is however possible to model $\Box$ by a monoidal
comonad $\D$, and positive types by comonad coalgebras with an action
$\texpose : P \to \D\,P$ that makes the services of the comonad
available at all value types, not just at $\Box N$.

Building on these observations, we investigate in this article how
CBPV %would look like if it would
can
keep track of not just the presence
of an effect or coeffect but also the \emph{kind} of effect and coeffect.
Semantically, precise information about effects can be obtained by
using a family of monads $\T[e]$ graded over effect classifiers $e$
drawn from a preordered monoid \cite{katsumata:popl14}.  For CBPV, we
generalize the concept of monad algebras $N$ to \emph{graded} such
ones, $N_e$.  Syntactically, computation typing $\CompTy \Gamma t e N$
is extended by an effect classifier $e$
(as usual in type and effect systems \citep{nielsen:effect99})
such that a computation can be
interpreted by a morphism $\Gamma \to N_e$.

Coeffect type systems
\citep{brunel:esop14,orchard:icfp14,ghicaSmith:esop14} have been
developed to track resource consumption by attaching usage information
to each variable in the typing context.  Coeffect typing generalizes
linear typing \citep{girard:linear} to quantitative typing
\citep{sergeyVytiniotisPeytonJones:popl14,mcBride:wadler60,atkey:lics18}
and subsumes sensitivity analysis \citep{reedPierce:icfp10} and static
information control flow \citep{volpano:jcs96} aka security typing
\citep{abadiBanerjeeHeintzeRiecke:popl99}.
Coeffects can be modeled by a comonad $\D[r]$ graded over resource
qualifiers $r$ drawn from a preordered semiring.  For the sake of
coeffect-graded CBPV, we generalize graded comonads to graded comonad
coalgebras.  Value types $P$ and contexts $\Gamma$ are then
interpreted as such coalgebras.  Value typing
$\qValTy \gamma \Gamma v P$ and computation typing
$\qCompTy \gamma \Gamma t N$ are based on linear typing and come
equipped with a resource context~$\gamma$.  Semantically, values are
interpreted as morphisms $\Gamma_{r\gamma} \to P_r$ allowing the
``multiplication'' or \emph{scaling} of a value by $r$ % natural in $r$
and computations as morphisms $\Gamma_\gamma \to N$.

Finally, effect and coeffect graded CBPV can be combined into a fully
graded CBPV calculus, surprisingly without sorting out any interaction
between effects and coeffects, such as the distributive laws of
\citet{orchard:icfp16}.  We credit the smoothness of the integration
to the careful placement of monad and comonad in CBPV's type system,
so that scaling is restricted to values and does not arise for computations.
% and the lack of need for scaling % of not just values but also
% of computations.

@@

Our style of effect tracking and its interpretation via graded monads
is based on \citet{wadlerThiemann:tocl03} and \citet{katsumata:popl14}.

@@

\citet{mcDermottMycroft:esop19} observe that CBPV can represent
call-by-name and call-by-value evaluation strategies but not
call-by-need.  As a remedy, they extend CBPV to ECBPV by variables
$\underline x : \diamond P$ that hold monadic values, plus a
corresponding let-binding construct.  They further present an
effect-graded version of ECBPV, drawing effect qualifiers from a
preordered monoid.  In contrast to our version, effect annotations
$f$ are stored at monadic types $\compty f P$.  Instead of our
judgement $\CompTy \Gamma t e N$ that accumulates effect traces $e$
and a semantic interpretation of computation types $N$ as graded monad
algebras, they define a syntactic operation $\compty f N$ that pushes
down effect qualifier $f$ to the monadic types $\compty {f'} P$ in $N$
where the effects are combined to $\compty{f \bu f'} P$.  This way of
handling effect grading seems equivalent to our approach, albeit it is
syntactic rather than semantic.

In previous work, \citet{mcDermottMycroft:ocs18} use coeffect typing
to track variable usage in call-by-need simply-typed lambda calculus
with conditionals.  Compared to the usual resource semiring $\R$
tailored to call-by-name usage analysis, their coeffect algebra is
more expressive: it can produce traces of variable usage, including
usage of bound variables.  They show how graded effect tracking can be
simulated by the coeffect typing, \eg, for the effect of
non-determinism.  Naturally, they focus on operational semantics.
